Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2019-12-21 article-information-2019.tex
commit submitted version

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents: adverse impact ratio (AIR), marginal effect (ME), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to present a template workflow for important ML applications that require high accuracy and interpretability and minimal discrimination.}

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute $X_j$, $c$ denotes the control group (often whites or males), $p$ indicates a protected group,  and $\text{Pr}(\hat{\mathbf{y}}|X_j)$ indicates a conditional probability regarding model outcomes, $\hat{\mathbf{y}}$, and some feature, $X_j$, evaluated over the relevant partition of data.  ME is a favored DI metric used by the CFPB, the primary agency charged with regulating fair lending laws at the largest US lending institutions and various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important measure of DI is adverse impact ratio (AIR), more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

Furthermore, using some recently developed discrimination mitigation methods may lead to non-compliance with anti-discrimination laws and regulations.  A fundamental maxim of US anti-discrimination law is that (to slightly paraphrase), ``similarly situated people should be treated similarly.''\footnote{As discussed in \textit{Bazemore v. Friday}, 478 US 385 (1986), a discrimination in pay case, the US Supreme Court wrote that,``Each week's paycheck that delivers less to a black than to a similarly situated white is a wrong actionable ...''  Beyond the obvious conceptual meaning, what specifically constitutes \textit{similarly situated} is controversial and its interpretation differs by circuit.} A model developed without inclusion of class status (or proxies thereof) usually considers similarly situated people the same on the dimensions included in the model: people who have the same feature values will usually have the same model output.  Obviously, the inclusion of protected class status will change model output by class.  With possible rare exceptions, this is likely to be viewed with legal and regulatory skepticism today, even if including class status is done with fairness as the goal.  Preprocessing and post-processing techniques may be similarly problematic, because industries that must provide explanations to those who receive unfavorable treatment (e.g., adverse action notices in US financial services) may have to incorporate the class adjustments into their explanations as well.

@@

The black-box nature of ML, the perpetuation or exacerbation of discrimination by ML, or the security vulnerabilities inherent in ML are each serious and difficult problems on their own. However, evidence is mounting that these harms can also manifest as complex intersectional challenges, e.g., the \textit{fairwashing} or \textit{scaffolding} of biased models with ML explanations, the privacy harms of ML explanations, or the adversarial poisoning of ML models to become discriminatory \cite{shokri2019privacy}, \cite{fair_washing}, \cite{scaffolding}.\footnote{See: \href{https://www.theguardian.com/technology/2016/mar/24/tay-microsofts-ai-chatbot-gets-a-crash-course-in-racism-from-twitter}{\textit{Tay, Microsoft's AI chatbot, gets a crash course in racism from Twitter}}.}\textsuperscript{,}\footnote{While the focus of this paper is not ML security, proposed best-practices from that field do point to transparency of ML systems as a mitigating factor for some ML attacks and hacks \cite{papernot2018marauder}. High system complexity is sometimes considered a mitigating influence as well \cite{hoare19811980}. This is sometimes known as the \textit{transparency paradox} in data privacy and security, and it likely applies to ML security as well, especially in the context of interpretable ML models and post-hoc explanation.\textsuperscript{25}}\textsuperscript{,}\footnote{See: \href{https://hbr.org/2019/12/the-ai-transparency-paradox}{\textit{The AI Transparency Paradox}}.} Again, this text makes no claims that the opacity, discrimination, or security problems in ML have been solved, even treated as independent problems. Instead, this text aims to highlight these issues as both singular entities and non-static intersectional phenomena. Practitioners should of course consider the discussed interpretable modeling, post-hoc explanation, and discrimination testing approaches as at least partial remedies to the black-box and discrimination issues in ML. However, they should also consider that explanations can ease model stealing, data extraction, and membership inference attacks, and that explanations can mask ML discrimination. Additionally, high-stakes, human-centered, or regulated ML systems should generally be built and tested with robustness to adversarial attacks as a primary design consideration, and specifically to prevent ML models from being poisoned or otherwise altered to become discriminatory. Accuracy, discrimination, and security characteristics of a system can change over time as well. Simply testing for these problems at training time, as presented in Sections \ref{sec:m_and_m} and \ref{sec:res}, is not adequate for high-stakes, human-centered, or regulated ML systems. Accuracy, discrimination, and security should be monitored in real-time and over time, as long as a model is deployed.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer of, and investor in, H2O.ai and a customer of BLDS, LLC. However, communications regarding XNN between Wells Fargo and the authors have been extremely limited prior to and during the drafting of this manuscript. Moreover, Wells Fargo exerted absolutely no editorial control over the text or results herein.}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This manuscript outlines a viable approach for training and evaluating machine learning (ML) systems for high-stakes, human-centered, or regulated applications using common Python programming tools. The accuracy and intrinsic interpretability of two types of constrained models, monotonic gradient boosting machines (MGBM) and explainable neural networks (XNN), a deep learning architecture well-suited for structured data, are assessed on simulated data with known feature importance and discrimination characteristics and on publicly available mortgage data. For maximum transparency and the potential generation of personalized adverse action notices, the constrained models are analyzed using post-hoc explanation techniques including plots of partial dependence (PD) and individual conditional expectation (ICE) and global and local Shapley feature importance. The constrained model predictions are also tested for disparate impact (DI) and other types of discrimination using measures with long-standing legal precedents: adverse impact ratio (AIR), marginal effect (ME), standardized mean difference (SMD), and additional straightforward group fairness measures. By combining interpretable models, post-hoc explanation, and discrimination testing with accessible software tools, this text aims to present a template workflow for important ML applications that require high accuracy and interpretability and that mitigate risks of discrimination.}

@@

\noindent where $X_p$ and $X_c$ represent binary markers created from some demographic attribute, $c$ denotes the control group (often whites or males), and $p$ indicates a protected group.  ME is a favored DI metric used by the CFPB, the primary agency charged with regulating fair lending laws at the largest US lending institutions and various other participants in the consumer financial market.\footnote{See: \href{https://files.consumerfinance.gov/f/201510_cfpb_supervisory-highlights.pdf}{Supervisory Highlights, Issue 9, Fall 2015}.} Another important measure of DI is adverse impact ratio (AIR), more commonly known as a \textit{relative risk ratio} in settings outside of regulatory compliance.

@@

Furthermore, using some recently developed discrimination mitigation methods may lead to non-compliance with anti-discrimination laws and regulations.  A fundamental maxim of US anti-discrimination law is that (to slightly paraphrase), ``similarly situated people should be treated similarly.''\footnote{In the pay discrimination case, \textit{Bazemore v. Friday}, 478 US 385 (1986), the US Supreme Court found that,``Each week's paycheck that delivers less to a black than to a similarly situated white is a wrong actionable ...''  Beyond the obvious conceptual meaning, what specifically constitutes \textit{similarly situated} is controversial and its interpretation differs by circuit.} A model developed without inclusion of class status (or proxies thereof) considers similarly situated people the same on the dimensions included in the model: people who have the same feature values will have the same model output (though there may be some small or random differences in outcomes due to computational issues). Obviously, the inclusion of protected class status will change model output by class. With possible rare exceptions, this is likely to be viewed with legal and regulatory skepticism today, even if including class status is done with fairness as the goal.\footnote{In a reverse discrimination case, \textit{Ricci v Desafano}, 557 US 557 (2009), the court found that any consideration of race which is not justified by correcting for past proven discrimination is illegal and, moreover, a lack of fairness is not necessarily evidence of illegal discrimination.}  Preprocessing and post-processing techniques may be similarly problematic, because industries that must provide explanations to those who receive unfavorable treatment (e.g., adverse action notices in US financial services) may have to incorporate the class adjustments into their explanations as well.

@@

The black-box nature of ML, the perpetuation or exacerbation of discrimination by ML, or the security vulnerabilities inherent in ML are each serious and difficult problems on their own. However, evidence is mounting that these harms can also manifest as complex intersectional challenges, e.g., the \textit{fairwashing} or \textit{scaffolding} of biased models with ML explanations, the privacy harms of ML explanations, or the adversarial poisoning of ML models to become discriminatory \cite{shokri2019privacy}, \cite{fair_washing}, \cite{scaffolding}.\footnote{See: \href{https://www.theguardian.com/technology/2016/mar/24/tay-microsofts-ai-chatbot-gets-a-crash-course-in-racism-from-twitter}{\textit{Tay, Microsoft's AI chatbot, gets a crash course in racism from Twitter}}.}\textsuperscript{,}\footnote{While the focus of this paper is not ML security, proposed best-practices from that field do point to transparency of ML systems as a mitigating factor for some ML attacks and hacks \cite{papernot2018marauder}. High system complexity is sometimes considered a mitigating influence as well \cite{hoare19811980}. This is sometimes known as the \textit{transparency paradox} in data privacy and security, and it likely applies to ML security as well, especially in the context of interpretable ML models and post-hoc explanation.\textsuperscript{26}}\textsuperscript{,}\footnote{See: \href{https://hbr.org/2019/12/the-ai-transparency-paradox}{\textit{The AI Transparency Paradox}}.} Again, this text makes no claims that the opacity, discrimination, or security problems in ML have been solved, even treated as independent problems. Instead, this text aims to highlight these issues as both singular entities and non-static intersectional phenomena. Practitioners should of course consider the discussed interpretable modeling, post-hoc explanation, and discrimination testing approaches as at least partial remedies to the black-box and discrimination issues in ML. However, they should also consider that explanations can ease model stealing, data extraction, and membership inference attacks, and that explanations can mask ML discrimination. Additionally, high-stakes, human-centered, or regulated ML systems should generally be built and tested with robustness to adversarial attacks as a primary design consideration, and specifically to prevent ML models from being poisoned or otherwise altered to become discriminatory. Accuracy, discrimination, and security characteristics of a system can change over time as well. Simply testing for these problems at training time, as presented in Sections \ref{sec:m_and_m} and \ref{sec:res}, is not adequate for high-stakes, human-centered, or regulated ML systems. Accuracy, discrimination, and security should be monitored in real-time and over time, as long as a model is deployed.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{XNN was first made public by the corporate model validation team at Wells Fargo bank. Wells Fargo is a customer of, and investor in, H2O.ai and a client of BLDS, LLC. However, communications regarding XNN between Wells Fargo and the authors have been extremely limited prior to and during the drafting of this manuscript. Moreover, Wells Fargo exerted absolutely no editorial control over the text or results herein.}
